---
title: "Project proposal"
author: "Energyyy"
output: github_document
---

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
#install.packages("lubridate")
library(lubridate)
library(tidyr)
```

## 1. Introduction

The United States of America has on of the most diverse electric grids in the world. With this many stakeholders it is interesting to know which factors affect electricity distribution and pricing to what extent in the US. We are looking into the data from the the U.S. Energy Information Administration (EIA). EIA collects, analyzes, and disseminates independent and impartial energy information. The utility companies are required to self-report this data due to regulations. The variables that we will be looking at are Ownership type, Customers (Count), Sales (Megawatt hours) Revenues (Thousands Dollars), and Average Price (cents/kWh).

Research Question: 
Which factors affect electricity distribution and pricing to what extent in the US?
- black outs
- urban rural
- sectors
-seasons
- regional differences
- utility companies ownership

## 2. Data
```{r readxl-library}
library(readxl)
```

```{r read-data}
table_Residential <- read_excel("/cloud/project/data/table_Residential.xlsx",
                                na = c("."))
table_Commercial <- read_excel("/cloud/project/data/table_Commercial.xlsx",
                               na = c("."))
table_Industrial <- read_excel("/cloud/project/data/table_Industrial.xlsx", 
                               na = c("."))
table_Transportation <- read_excel("/cloud/project/data/table_Transportation.xlsx", 
                                   na = c("."))
table_Disturbance <- read_excel("/cloud/project/data/table_Disturbance.xlsx",
                                na= c(".", ". Hours,  . Minutes", "Unknow", ".        ."))

```

```{r glimpse-data}
glimpse(table_Residential)
glimpse(table_Commercial)
glimpse(table_Industrial)
glimpse(table_Transportation)
glimpse(table_Disturbance)
```

This mutate function adds a column which contains the sector of the data.
```{r mutate-to-sector}
residential <- table_Residential %>%
  mutate(Sector = c("Residential"))
commercial <- table_Commercial %>%
  mutate(Sector = c("Commercial"))
transportation <- table_Transportation %>%
  mutate(Sector = c("Transportation"))
industrial <- table_Industrial %>%
  mutate(Sector = c("Industrial"))
```

This code binds all the tables with sectors together. 
```{r bind-rows}
energy_sector <- bind_rows(residential, commercial, industrial, transportation)
glimpse(energy_sector)
```

```{r separate-duration-disturbance}
disturbance_unjoined <- table_Disturbance |>
  separate(col = Duration, into = c("Hours (Duration)", "Minutes (Duration)"), sep = ',') 

disturbance <- 
  rename(disturbance_unjoined, Entity = `Utility/Power Pool`) |>
  right_join(energy_sector, by = "Entity")

glimpse(disturbance)
```


## 3. Ethics review

### Limitations in data sources:
- table_CAIDI: Most utility companies use IEE standards to report CAIDI, but some do not.  We have decided to leave out companies of our analysis that do not use IEE standards, so we will not have a complete account of ALL utility companies, but of most of them. We decided not to use the CAIDI index that includes Major Event Days because those represent hurricanes, storms etc. which utility companies do not have control over. Thus, they do not attest to the utility's ability to provide reliable services.

- The data collection is standardized and done by governmental institutions, so the data should be reliable, of high quality and equal across utility companies and sectors. However, the EIA relies on the honesty of the utility company to share accurate data with the EIA. 

- Ilham and Anna do not have an extensive background in the energy industry, so we rely on Rudy for most of the expertise in our project.

### Positive effects on people:
- Education of lay people: might help citizens to be better informed of the US energy industry which is highly complex.
- Decision-makers: gain a better understanding to improve policy-making and can therefore evaluate the efficiency and effectiveness of utility companies. This enables them to encourage the companies to become more efficient and effective. 
- Education of Rudy: improves his personal understanding of the US energy sector
- We could enhance these positive effects by dispersing our research more and communicating it to citizens and decision-makers.

### Negative effects on people:
- The reputation of specific utility companies might suffer from our research if they are connected to higher prices for lower-quality offering of energy services.

### Minimising negative impact:
- We will talk about the limitations/disclaimers of our research, so that people are aware of what we cannot show with our data, so that people do not draw wrong conclusions from our research. 
  - This is a beneficial action to take because people will be more aware of the limitations and can understand our data and our visualizations better. 
- We are presenting to our class at COA only and we do not intend to share our outcomes with the wider community where it would lead to the implementation of policies etc.

## 4. Data analysis plan

We will use all the mentioned variables in our analyses. 
There will be other data that we include in our analysis to visualize the data we already have, such as coordinates to locate the US states on a potential heatmap. In case we would like to expand our data analysis, we would go further into price developments per sector and US state potentially using data between 1960-2023. 

Priority list of data analyses:

1. Ownership vs price of electricity /according to sector 
- Q: Does ownership e.g. cooperatives influence the average price of electricity, does it benefit individuals/residential sector or is it more geared towards commercial sector…?`
```{r ownership-average-price-state}
p1 <- energy_sector |>
  filter(!is.na(Ownership)) |>

 ggplot(aes(x = Ownership, y = `Average Price (cents/kWh)`)) +
  #geom_point(aes(color = Sector)) +
  geom_boxplot(outlier.shape = NA, aes(color = Sector)) +
  ylim(0,25)+
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust  = 1), plot.margin = margin(1, 0.1, 1, 1, unit = "mm")) 
#ggsave(filename = "plot1.png", device = "png")
```

2. Revenue vs price and actual blackouts and ownership 
  - Q: does less profit and community owned mean less blackouts? Or does ownership tell us that big companies have more means to prevent blackouts?

```{r count-disturbances-per-companies}
disturbance |> 
  group_by(Entity) |>
  count() |>
  arrange(n)
```

For the next plot we want to count the disturbances per utility company. However, we need to clean up the data first because some companies come up with different names. Hence, the mutate function. 
```{r mutate-to-group-same-companies-different-names}
disturbance |>
  mutate(Company = 
           case_when(
    `Entity` %in% 
      c("American Electric Power (Regulated Generation)", 
      "American Electric Power - (RFC Reliability Region) (8400 Smiths Mill Road, New Albany Ohio 43054)")
        ~ "American Electric Power", 
     `Entity` %in% 
      c("Baltimore Gas & Electric Co", 	
"Baltimore Gas and Electric") ~ "Baltimore Gas & Electric",
     TRUE ~ as.character(as.character(`Entity`))
    )) |>
    group_by(Company) |>
  count()
  
#ggplot(aes(x = `Average Price (cents/kWh)`, y = `Revenues (Thousands Dollars)`, size = n of blackouts/in categories)), facet(~ ownership)
```
  

3. Amount of electricity used, cheaper prices?
  - Q: Is it cheaper to buy electricity depending on the sector and is buying electricity in bulk cheaper? 
  - ggplot(aes(x = Sales(mWh)/n of customers, y = average price, color = sectors))

4. Number of customers vs price by sector 
  - Q: Do bigger utility companies (meaning more customers) offer lower avg. prices due to scale?
  - ggplot(aes(x = avg price, y = customers)), facet( ~ sector)

5. Per household consumption of electricity (megawatthours sold/customers) vs price 
  - Q: Do consumers use less electricity if it is more expensive? 
  - ggplot(aes(x = Sales/customers, y = avg. price))

Stretch goals: 
6. Heatmap of the US' electricity use
  - Note: we might need long lat data to map the US states
  - Average per capita Sales per state (sum up all sales, group by state, calculate average), each box of the heatmap is one state

7. Seasonal price variation vs average temperature, per month
  - We would need: avg temperature per month per state, electricity price per month for multiple years per state
