---
title: "Project Workspace"
author: "Energyyy"
output: github_document
---

```{r knitr}
library(knitr)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# 1. Load packages

```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
library(tidyr)
library(readxl)
#install.packages("leaflet")
library(leaflet)
#install.packages("sf")
library(sf)
#install.packages("gganimate")
library(gganimate)
library(ggthemes)
library(gapminder)
#install.packages("visdat")
library(visdat)
#install.packages("naniar")
library(naniar)
```

# 2. Introduction
The United States of America has one of the most diverse electric grids in the world. With that many stakeholders it is interesting to know which factors affect electricity distribution and pricing to which extent in the US. We are looking into data from the U.S. Energy Information Administration (EIA). EIA collects, analyzes, and disseminates independent and impartial information on the US energy sector. The utility companies are required to self-report this data due to national regulations. The variables that we will be looking at are ownership type, customers (Count), sales (Megawatt hours), revenues (Thousands Dollars), and average price (cents/kWh), US state, energy sectors, disturbances, and CAIDI - a reliability index.

## Research Question 
Which factors affect electricity distribution and pricing to what extent in the US?

- sectors

- regional differences in reliability

- utility company's ownership

# 3. Data
```{r read-data}
table_Residential <- read_excel("/cloud/project/data/table_Residential.xlsx",
                                na = c("."))
table_Commercial <- read_excel("/cloud/project/data/table_Commercial.xlsx",
                               na = c("."))
table_Industrial <- read_excel("/cloud/project/data/table_Industrial.xlsx", 
                               na = c("."))
table_Transportation <- read_excel("/cloud/project/data/table_Transportation.xlsx", 
                                   na = c("."))
table_Disturbance <- read_excel("/cloud/project/data/table_Disturbance.xlsx",
                                na= c(".", ". Hours,  . Minutes", "Unknow", ".        ."))
table_CAIDI <- read_excel("/cloud/project/data/table_CAIDI.xlsx")

USA_df <- sf::st_read("/cloud/project/data/USA_States_Generalized.shp")
```

# 4. Ethics review

## Limitations in data sources:
- table_CAIDI: Most utility companies use IEE standards to report CAIDI, but some do not.  We have decided to leave out companies of our analysis that do not use IEE standards, so we will not have a complete account of ALL utility companies, but of most of them. We decided not to use the CAIDI index that includes Major Event Days because those represent hurricanes, storms etc. which utility companies do not have control over. Thus, they do not attest to the utility's ability to provide reliable services.

- The data collection is standardized and done by governmental institutions, so the data should be reliable, of high quality and equal across utility companies and sectors. However, the EIA relies on the honesty of the utility company to share accurate data with the EIA. 

- Ilham and Anna do not have an extensive background in the energy industry, so we rely on Rudy for most of the expertise in our project.

## Positive effects on people:
- Education of lay people: might help citizens to be better informed of the US energy industry which is highly complex.
- Decision-makers: gain a better understanding to improve policy-making and can therefore evaluate the efficiency and effectiveness of utility companies. This enables them to encourage the companies to become more efficient and effective. 
- Education of Rudy: improves his personal understanding of the US energy sector
- We could enhance these positive effects by dispersing our research more and communicating it to citizens and decision-makers.

## Negative effects on people:
- The reputation of specific utility companies might suffer from our research if they are connected to higher prices for lower-quality offering of energy services.

## Minimising negative impact:
- We will talk about the limitations/disclaimers of our research, so that people are aware of what we cannot show with our data, so that people do not draw wrong conclusions from our research. 
  - This is a beneficial action to take because people will be more aware of the limitations and can understand our data and our visualizations better. 
- We are presenting to our class at COA only and we do not intend to share our outcomes with the wider community where it would lead to the implementation of policies etc.

# 5. Tidying data

This mutate function adds a column which contains the sector of the data.
```{r mutate-to-sector}
USA_df <- USA_df %>%
 rename(`Census Division\r\nand State` = STATE_NAME)
residential <- table_Residential %>%
  mutate(Sector = c("Residential"))
commercial <- table_Commercial %>%
  mutate(Sector = c("Commercial"))
transportation <- table_Transportation %>%
  mutate(Sector = c("Transportation"))
industrial <- table_Industrial %>%
  mutate(Sector = c("Industrial"))
```

This code binds all the tables with sectors together. 
```{r bind-rows}
energy_sector <- bind_rows(residential, commercial, industrial, transportation)
```

```{r table-disturbance-preparation}
table_Disturbance <- table_Disturbance |>
  mutate(Entity = `Utility/Power Pool`)

energy_sector_rev_sum <- energy_sector |> 
  select(Entity, Ownership, `Revenues (Thousands Dollars)`) |>
  group_by(Entity, Ownership) |>
  summarise(sum_revenue = sum(`Revenues (Thousands Dollars)`)) |>
  as.data.frame()

energy_sector_customer_sum <- energy_sector |> 
  select(Entity, Ownership, `Customers (Count)`) |>
  group_by(Entity, Ownership) |>
  summarise(sum_customer = sum(`Customers (Count)`)) |>
  as.data.frame()
```

```{r pivot-CAIDI}
table_CAIDI <- table_CAIDI |>
 pivot_longer(cols = `2013`:`2022`,
              names_to = "Year",
              values_to = "CAIDI")
```

```{r count-disturbances-per-companies-revenue}
disturbance_count <- table_Disturbance |>
  group_by(Entity) |>
  count()

disturbance_complete <- disturbance_count |> 
  inner_join(energy_sector_rev_sum)
```

```{r count-disturbances-per-companies-ownership}
disturbance_ownership <- disturbance_count |> 
  inner_join(energy_sector_customer_sum)
```


## Missing data
```{r vis-missing-data, fig.alt = "Graph showing percentage of missing data points for each variable in three dataframes (Energy sector, Disturbance, CAIDI). Most datasets are almost complete with the Disturbance dataset missing the most data points at 3.9 %"}
vis_miss(energy_sector)

vis_miss(table_Disturbance)

vis_miss(table_CAIDI)
```

## Theme
```{r make-our-own-theme}
our_theme <- theme_grey() + 
  theme(text = element_text(color = "#1d91c0"), 
        legend.text = element_text(color = "black"))
  
```

# 6. Data Analysis

## 6.1. Ownership vs price of electricity 
**Q:** Does ownership affect the price of electricity? We wonder whether  cooperatives offer on average a lower price of electricity because that is what the proposers of the Pine Tree Power bill in Maine have recently advertised to convince people to vote in favor of buying out investor-owned company to turn them into cooperatives. And, looking at differences between sectors, is there a general trend for all sectors or do different ownership types offer different price levels to different sectors? E.g. Are cooperatives more beneficial for the residential sector? 

```{r ownership-average-price-state, fig.alt="Boxplot with trendline of average price in cents per kilowatthour in 2022 in each sector according to type of ownership. Types of ownership are ordered on a spectrum from mor cooperative to more investor-owned. Federally-owned companies and companies owned by political subdivisions seem to offer lower than average prices in all sectors. The industrial sector generally has lower average prices than the other sectors."}
energy_sector |>
  filter(!is.na(Ownership)) |>
  mutate(
    Ownership = fct_relevel(
      Ownership,
      "Cooperative", "Municipal", "Political Subdivision", "State", "Federal", "Behind the Meter", "Retail Power Marketer", "Investor Owned"
    )) |>

 ggplot(aes(x = as.numeric(Ownership), y = `Average Price (cents/kWh)`)) +
  geom_boxplot(outlier.shape = NA, aes(color = Ownership)) +
  geom_smooth(method = "loess", color = "black") +
  ylim(0,25)+
  facet_wrap(~Sector) +
  scale_color_manual(values = c("Cooperative" = "#ffffd9", "Municipal" = "#edf8b1", "Political Subdivision" = "#c7e9b4", "State" = "#7fcdbb", "Federal" = "#41b6c4", "Behind the Meter" = "#1d91c0", "Retail Power Marketer" = "#225ea8", "Investor Owned" = "#0c2c84")) +
  labs(
    title = "Average price (cents/kWh) according to type of ownership",
    subtitle = "Per sector based on 2022 data",
    caption = "Source : https://www.eia.gov/electricity/data.php#sales",
    x = "") +
  our_theme +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        panel.grid = element_blank())
#ggsave(filename = "Ownership vs price.png", device = "png")
```

**A:** Federally-owned utility companies offer below average prices in all sectors. All other ownerships do not deviate significantly from the average price. We do not pay a lot of attention to the transportation sector, even though its trendline looks different and interesting, because there are only 43 data points in comparison to more than 1500 data entries for the commercial and residential sector.

```{r vis-number-of-data-points-sector, fig.alt = "Barplot showing number of data points for each sector. Number ranges from 1500 entries for the commercial and residential sectors to 43 entries for the transportation sector."}
ggplot(energy_sector, aes(x = forcats::fct_infreq(Sector), fill = Sector)) +
  geom_bar() +
  labs(title = "Number of data points per sector", x = "Sector", y = "Count") +
  scale_fill_viridis_d() +
  our_theme +
  theme(legend.position = "none") 
```

## 6.2. Disturbance and Reliability 

### 6.2.1. Blackouts and ownership 
**Q:** How do blackouts, revenue and ownership relate to each other? Do cooperatives that are less profit-oriented, but customer-oriented correlate with less blackouts? Or do bigger companies have more means to prevent blackouts than small cooperatives, even if they are for profit and investor-owned?

```{r disturbances-vs-revenue-and-ownership, fig.alt="Scatterplot of number of Disturbances per utility company in relation to the type of ownership of the utility company and the revenue made by the company in 2022 in Millions of dollars. Size of the points reflects number of disturbances experienced in 2022. Investor-owned companies experience by far the most disturbances."}
ggplot(disturbance_complete, aes(
  y = sum_revenue/1000000, 
  x = `Ownership`,
  size = n)) +
  geom_point(color = "#1d91c0", alpha = 0.8) +
  labs(title = "Number of disturbances per utility company", 
       subtitle = "in relation to ownership type and revenue", 
       x = "Type of ownership", y = "Revenue in Million Dollars ($)", 
       size = "Number of\ndisturbances",
       caption = "Source : https://www.eia.gov/electricity/data.php#sales") +
  our_theme
```

**A:** Most disturbances occur in investor-owned companies. Revenue does not seem to play a major role in how many disturbances occur as the disturbances in investor-owned companies happen in all companies with low and high revenues. To check whether the number of disturbances is so high for investor-owned companies because there are simply more investor-owned companies than e.g. cooperatives or municipally-owned utilities, we plotted the number of data entries for each type of ownership.  

```{r vis-number-of-data-points-ownership, fig.alt = "Barplot showing number of data points for each Type of Ownership. Number ranges from 1800 entries for cooperatives and 600 for investor-owned companies to around 50 entries for state-owned companies."}
energy_sector_customer_sum |>
  filter(!is.na(Ownership)) |>
ggplot(aes(x = forcats::fct_infreq(Ownership), fill = Ownership)) +
  geom_bar() +
  labs(title = "Number of data points per type of ownership", x = "Type of ownership", y = "Count", caption = "Source : https://www.eia.gov/electricity/data.php#sales") +
  scale_fill_viridis_d() +
  our_theme +
  theme(legend.position = "none") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
This shows us that the number of disturbances in investor-owned companies is not high simply because there are many companies with this type of ownership because there are three times as many cooperatives and they still do not have a lot of disturbances. 

**Q:** Maybe, most disturbances are related to investor-owned companies because they simply serve more customers?

```{r number-of-customers-per-ownership, fig.alt = "Boxplot showing number of customers of each utility company versus ownership. Investor-owned companies have by far the most customers with an average of 500000 customers per utility company, some even serving more than 4 million households. All other ownership types serve less than 100000 customers on average."}
energy_sector_customer_sum |>
  filter(!is.na(Ownership)) |>
ggplot(aes(x = Ownership, y = sum_customer/1000000)) +
  geom_boxplot() +
  ylim(0, 4) +
  labs(title = "Number of customers versus ownership", subtitle = "in each utility company", y = "Number of customers (in Millions)") +
  our_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**A:** Investor-owned companies have by far the most customers. All other ownership types serve a lot less customers on average. Thus, the amount of disturbances in investor-owned companies might correlate positively with the number of customers they serve. 

```{r disturbances-customers-average-per-ownership, fig.alt = "Scatterplot with trendline showing average number of customers versus average number of disturbances per type of ownership. Investor-owned companies have a below-average number of disturbances, while companies owned by political subdivisions have an above average number of disturbances."}
disturbance_ownership_average <- disturbance_ownership |>
  group_by(Ownership) |>
  summarise(mean(sum_customer), mean(n))
  
ggplot(disturbance_ownership_average, aes(
         x = `mean(sum_customer)`/1000, 
         y = `mean(n)`)) +
   geom_smooth(method = "glm", alpha = 0.5) +
  geom_point(aes(color = Ownership), size = 3) +
  scale_color_viridis_d() +
  labs(title = "Average number of customers vs number of disturbances",
       subtitle = "per type of ownership", 
       x = "Average number of customers (in Thousands)", 
       y = "Average number of disturbances") +
  our_theme

#ggsave(filename = "Average_customer_disturbances.png", device = "png")
```

**A:** There seems to be a positive correlation between number of customers per type of ownership and number of disturbances. Investor-owned companies have a below-average number of disturbances, while companies owned by political subdivisions have an above average number of disturbances. Federally-owned and cooperative-owned utilities have low average numbers of customers and low numbers of disturbances.
  
### 6.2.2. Electricity network reliability across the US 

**Q:** Not taking into account major event days, such as hurricanes, storms, etc. that cause blackouts, how does the Customer Average Interruption Duration Index (CAIDI) vary throughout the US states over time from 2013-2022? 

```{r animate-gif-CAIDI, eval=FALSE, fig.alt = "Map showing value of of the Customer Average Interruption Duration Index including data from 2013 to 2022 for the lower 48 of the US states. This data does not include major events such as natural catastrophes. The index ranges from 0 to 250. West Virginia and Michigan are consistently suffering from low reliability in electricity supply while Florida and North Dakota have a low number of interruptions."}
# Create a base plot
CAIDI_leaflet_df$Year <-as.numeric(CAIDI_leaflet_df$Year)

p <- CAIDI_leaflet_df |>
  ggplot(aes(fill = CAIDI,
             group = Year)) +
  geom_sf() +
  xlim(125, 68) +
  ylim(24.5, 50) +
  scale_fill_viridis_c() +
  our_theme +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), 
        axis.ticks.y = element_blank(),
        panel.grid = element_blank(),
        panel.background = element_rect(fill='white')) +
  labs(title = "Customer Average Interruption Duration Index (CAIDI) map of lower 48 US states", 
       subtitle = "Year: {closest_state}",
       caption = "Source: https://www.eia.gov/electricity/data/annual/") 

# Create an animated plot
p <- p + transition_states(
  states = Year,
  transition_length = 3,
  state_length = 3)
#animate(p, duration = 30, renderer = gifski_renderer("map.gif"))
```

**A:** West Virginia and Michigan are consistently suffering from low reliability in electricity supply while Florida and North Dakota have a low number of interruptions. There are no dramatic changes for any of the states over time. 

## 6.3. Amount of electricity consumed vs average price

**Q:** How does the amount of electricity consumed per customer affect price? Is it cheaper to buy electricity depending on the sector? 

Historically, the industrial sector benefited from lower electricity prices because they constituted a reliable demand and were the preferred customers for utility companies because it was cheaper for the utility company to provide electricity to customers with a constant, reliable demand than it was for them to provide to the residential sector where demand always varies (Bakke, 2016).

Does this tendency toward the industrial sector still exist today?

Bakke, G., 2016. "The Grid: The Fraying Wires Between Americans and Our Energy Future". Bloomsbury, USA. 

```{r average-price-in-kWH-consumption, fig.alt="Scatterplot showing average kilowatthour consumption per customer versus average price in cents per kilowatthour, faceted by sector. Each facet includes the mean average price offered by the companies to each sector. The industrial and transportation sector have the widest spread of electricity use while the Residential and Commercial sector use less electricity per customer. The industrial sector pays on average the least for their electricity with 9.25 cents per kilowatthour in comparison to 13.55 cents per kilowatthour in the residential sector where there are many customers which consume smaller amounts of electricity each."}
#creating the mean of average price in each sector
line_test <- energy_sector |>
  summarise(mean(`Average Price (cents/kWh)`, na.rm = TRUE),.by = Sector)

line_test$mean <- line_test$`mean(\`Average Price (cents/kWh)\`, na.rm = TRUE)` 

line_test$label <-
  round(line_test$mean, 2)

#create graph using mean of average price in each sector
energy_sector |>
  ggplot(aes(x = (
    `Sales (Megawatthours)`*1000/`Customers (Count)`), y = `Average Price (cents/kWh)`, color = Sector)) +
  geom_point() +
  geom_smooth() +
  geom_hline(data = line_test, aes(yintercept = mean)) +
  geom_text(data = line_test, aes(50000000, y = label, label = label, vjust = -1), color = "black") +
  facet_wrap(~Sector) +
  scale_color_viridis_d() +
  xlim(0, 400000000) +
  labs(
    title = "Average kWh consumption per customer vs average price",
    subtitle = "by sector in 2022",
    x = "Average kWh consumption per customer",
    y = "Average price (cents/kWh)",
    caption = "Source : https://www.eia.gov/electricity/data.php#sales"
  ) +
  our_theme +
  theme(panel.grid = element_blank())

ggsave(filename = "price_per_cust_consumption.png", device = "png")
```

**Q:** Yes, this tendency still exists today as the industrial sector with the highest electricity consumption per consumer still gets the lowest average prices. There does not seem to be a general trend of lower prices for higher electricity consumption per customer as none of the facets show a negative trendline.

## 6.4. Number of customers vs price in the residential sector

**Q:** Considering only the residential sector, how does the number of customers that a utility company serves affect the average price they offer? Are bigger utility companies with more customers able to offer lower average prices due to scale?
  
```{r price-by-customers, fig.alt = "Scatterplot and trendline of Number of Customers in the Residential Sector per Utility Company versus average price in cents per kilowatthour. There is no correlation between the number of customers and the average price. The price that the utility companies offer depends on other factors. Most utility companies serve electricity to less than 50000 households, while some serve 250000 and more households."}
energy_sector |>
  filter(Sector == "Residential") |>
  ggplot(aes(x = `Customers (Count)`, y = `Average Price (cents/kWh)`)) +
  geom_point() +
  geom_smooth() +
  xlim(0, 250000) +
  labs(title = "Average price vs. number of customers", 
       subtitle = "per utility company for the residential sector in 2022", 
       x = "Number of customers", 
       y = "Average price (cents/kWh)",
       caption = "Source : https://www.eia.gov/electricity/data.php#sales") +
  our_theme
```

**A:** Scale does not seem to play an important role in determining prices as average price does not lower with an increase in the number of customers. In fact, there is a slight opposite trend. 

## 6.5. Per household consumption of electricity vs price 

**Q:** Considering only the residential sector, how does electricity consumption vary in relation to the average price? Do consumers use less electricity if it is more expensive? 

This analysis does not consider general trends in electricity prices over time, but only the difference between different utility companies.
  
```{r household-consumption-electricity, fig.alt="Heatmap and trendline of average electricity consumption per customer in the Residential Sector in kilowatthours of each utility company versus average price in cents per kilowatthours offered by the respective utility company. Heatmap shows that most customers use around 14000 kilowatthours of electrcity annually paying between 10 to 14 cents per kilowatthour. There is a slight negative correlation. Household consumption is lower when the price is high."}
ggplot(table_Residential, aes(x = (`Sales (Megawatthours)`*1000)/`Customers (Count)`, y = `Average Price (cents/kWh)`)) +
  geom_hex() +
  geom_smooth() +
  labs(title = "Electricity Consumption per Customer vs. Average Price", 
       subtitle = "in the Residential Sector in 2022", 
       x = "Average Electricity Consumption per Customer (in kWh)", 
       y = "Average Price (cents/kWh)", 
       fill = "Number of\nUtility\nCompanies",
       caption = "Source : https://www.eia.gov/electricity/data.php#sales") +
  our_theme

ggsave(filename = "heatmap_consumption.png", device = "png")
```

**A:** There is a slight negative correlation between average electricity consumption per customer and average price. Most customers use around 14,000 kilowatthours of electricity annually paying between 10 to 14 cents per kilowatthour. The per household consumption of electricity is slightly lower when the price is high.